--- 
title: "TMA4250 Spatial Statistics Exercise 1, Spring 2019"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
  pdf_document: default
  bookdown::html_document2: default
  html_document:
    df_print: paged
subtitle: 'Group members: Henrik Syversveen Lie, Øyvind Auestad'
references:
- id: Omre
  title: Bayesian Spatial Inversion with Conjugate Prior Models
  author:
  - family: Omre
    given: Henning
  type: article-journal
  issued:
    year: 2019
    month: 1
---


```{r setup, include = FALSE}
library(formatR)
showsol <- FALSE
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 68), tidy = TRUE, warning = FALSE, error = FALSE, message = FALSE, echo = TRUE)
```


```{r, echo = F, eval = T}
library(reshape2)
library(geoR)
library(ggplot2)
library(MASS)
library(cowplot)
library(fields)
library(akima)
library(spatial)
```

# Problem 1: Analysis of Point Patterns
## a)
We consider the three real data point patterns found in the R package `MASS`.
```{r, echo = F, eval = T, out.width = "33%"}
library(MASS)
data_cells <- ppinit("cells.dat")
data_redwood <- ppinit("redwood.dat")
data_redwood$y <- -data_redwood$y
data_pines <- ppinit("pines.dat")

plot(data_cells, main = "Data from cells.dat file", xlab = "x", ylab = "y")
plot(data_redwood, main = "Data from redwood.dat file",xlab = "x", ylab = "y")
plot(data_pines, main = "Data from pines.dat file",xlab = "x", ylab = "y")
```

## b)
We then want to compute the empirical $L$-function for each of the point patterns, and use the R function `Kfn()` to do this. The $L$-functions are then displayed for each point pattern.

```{r, echo = F, eval = T, out.width = "33%"}
cells_L <- Kfn(data_cells,1)
#cells_L$y <- cells_L$y/cells_L$x
redwood_L <- Kfn(data_redwood, 1)
#redwood_L$y <- redwood_L$y/redwood_L$x
pines_L <- Kfn(data_pines,7)
#pines_L$y <- pines_L$y/pines_L$x
plot(cells_L, type="s", xlab="distance", ylab="L(t)", main = "L-function for cells data")

plot(redwood_L, type="s", xlab="distance", ylab="L(t)", main = "L-function for redwood data")

plot(pines_L, type="s", xlab="distance", ylab="L(t)", main = "L-function for pines data")
```
The plots of the $L$-functions are closely related to the displayed point patterns. The cells data has a low value of the $L$-function for low values of $t$. This means that there is a low probability of points being close to each other, we have a repulsive model, which is verified on the displayed point pattern. The redwood data on the other hand, has a high value of the $L$-function for low values of $t$. This means that there is a large probability of points being close, we have a clustered model, which also is verified on the displayed point pattern. Finally we have the pines data, for which the $L$-function closely resembles a straight line. This means that the point pattern 

## c)
```{r, echo = F, eval = T, out.width = "33%"}
k_cells <- length(data_cells$y)
k_redwood <- length(data_redwood$y)
k_pines <- length(data_pines$y)
L_matrix_cells <- matrix(NA, nrow = 100, ncol = length(cells_L$x))
L_matrix_redwood <- matrix(NA, nrow = 100, ncol = length(redwood_L$x))
L_matrix_pines <- matrix(NA, nrow = 100, ncol = length(pines_L$x))
for (i in 1:100){
  df_cells = data.frame(x = runif(k_cells,0,1), y = runif(k_cells,0,1))
  df_redwood = data.frame(x = runif(k_redwood,0,1), y = runif(k_redwood,0,1))
  df_pines = data.frame(x = runif(k_pines,0,9), y = runif(k_pines,0,9))
  L_matrix_cells[i,] = Kfn(df_cells,1)$y
  L_matrix_redwood[i,] = Kfn(df_redwood,1)$y
  L_matrix_pines[i,] = Kfn(df_pines,7)$y
}

pred_interval_cells <- apply(L_matrix_cells,2,sort)
rownames(L_matrix_cells) = paste("trial", seq(100), sep="")
colnames(L_matrix_cells) = paste("x", seq(length(cells_L$x)), sep="")

dat_cells = as.data.frame(L_matrix_cells)
dat_cells$trial = rownames(L_matrix_cells)
mdat_cells = melt(dat_cells, id.vars="trial")
mdat_cells$x = as.numeric(gsub("x", "", mdat_cells$variable))/100
df_cells = data.frame(x = seq(0,1,length.out = 100), pred = cells_L$y, lb = pred_interval_cells[5,], ub = pred_interval_cells[95,])

ggplot() +
    theme_bw() +
    theme(panel.grid=element_blank()) +
    geom_line(data = mdat_cells, aes(x=x, y=value, group=trial), size=0.2, alpha=0.1) +
    geom_line(data = df_cells, aes(x = x, y = pred), size = 1, col = 'red') + 
    geom_line(data = df_cells, aes(x = x, y = lb), size = 1, col = 'blue', linetype = "dashed") + 
    geom_line(data = df_cells, aes(x = x, y = ub), size = 1, col = 'blue', linetype = "dashed") + ylab("L(t)") + xlab("t") + ggtitle("Cells")

pred_interval_redwood <- apply(L_matrix_redwood,2,sort)
rownames(L_matrix_redwood) = paste("trial", seq(100), sep="")
colnames(L_matrix_redwood) = paste("x", seq(length(redwood_L$x)), sep="")

dat_redwood = as.data.frame(L_matrix_redwood)
dat_redwood$trial = rownames(L_matrix_redwood)
mdat_redwood = melt(dat_redwood, id.vars="trial")
mdat_redwood$x = as.numeric(gsub("x", "", mdat_redwood$variable))/100
df_redwood = data.frame(x = seq(0,1,length.out = 100), pred = redwood_L$y, lb = pred_interval_redwood[5,], ub = pred_interval_redwood[95,])

ggplot() +
    theme_bw() +
    theme(panel.grid=element_blank()) +
    geom_line(data = mdat_redwood, aes(x=x, y=value, group=trial), size=0.2, alpha=0.1) +
    geom_line(data = df_redwood, aes(x = x, y = pred), size = 1, col = 'red') + 
    geom_line(data = df_redwood, aes(x = x, y = lb), size = 1, col = 'blue', linetype = "dashed") + 
    geom_line(data = df_redwood, aes(x = x, y = ub), size = 1, col = 'blue', linetype = "dashed") + ylab("L(t)") + xlab("t") + ggtitle("Redwood")

pred_interval_pines <- apply(L_matrix_pines,2,sort)
rownames(L_matrix_pines) = paste("trial", seq(100), sep="")
colnames(L_matrix_pines) = paste("x", seq(length(pines_L$x)), sep="")

dat_pines = as.data.frame(L_matrix_pines)
dat_pines$trial = rownames(L_matrix_pines)
mdat_pines = melt(dat_pines, id.vars="trial")
mdat_pines$x = as.numeric(gsub("x", "", mdat_pines$variable))*7/100
df_pines = data.frame(x = seq(0,7,length.out = 99), pred = pines_L$y, lb = pred_interval_pines[5,], ub = pred_interval_pines[95,])

ggplot() +
    theme_bw() +
    theme(panel.grid=element_blank()) +
    geom_line(data = mdat_pines, aes(x=x, y=value, group=trial), size=0.2, alpha=0.1) +
    geom_line(data = df_pines, aes(x = x, y = pred), size = 1, col = 'red') + 
    geom_line(data = df_pines, aes(x = x, y = lb), size = 1, col = 'blue', linetype = "dashed") + 
    geom_line(data = df_pines, aes(x = x, y = ub), size = 1, col = 'blue', linetype = "dashed") + ylab("L(t)") + xlab("t") + ggtitle("Pines")
```



# Problem 2: Bayesian inversion in Poisson RF
We consider an area of size $(300\times 300)$ m$^2$ containing a pine tree forest, and the actual locations of the pine trees are to be assessed. The pine tree locations are observed from a satellite by remote sensing, and due to partly cloudy weather the observation probability for individual trees vary across the area. We discretize the area into a regular $(30\times30)$-grid $L$ with grid unit size $\Delta_n = 100$m$^2$. The true, but unknown, number of pine trees located in each grid unit is the $n$-vector $k$, with $\{k(\mathbf{x}); \mathbf{x}\in L\}$. The probabilites for observing a tree in each grid unit is represented by the $n$-vector $\{\alpha(\mathbf{x});\mathbf{x}\in L\}$, which vary across the area. We also denote by the $n$-vector $\{d(\mathbf{x});\mathbf{x}\in L\}$ the number of trees observed in each grid unit.

## a)
First, we display the observations and the observation probabilities across the grid. The observation probabilities vary from $0.04$ to $0.67$, and the number of observed trees in each grid unit varies between 0 and 3.
```{r, echo = F, eval = T, out.width = "50%"}
obspines <- read.table("https://www.math.ntnu.no/emner/TMA4250/2018v/Exercise2/obspines.txt")[2:901,]
obsprob <- read.table("https://www.math.ntnu.no/emner/TMA4250/2018v/Exercise2/obsprob.txt")[2:901,]

colnames(obsprob) <- c("x","y","z")
colnames(obspines) <- c("x","y","z")

obspines$x <- as.numeric(obspines$x)
obspines$y <- as.numeric(obspines$y)
obspines$z <- as.numeric(obspines$z)

obsprob$x <- as.numeric(obsprob$x)
obsprob$y <- as.numeric(obsprob$y)
obsprob$z <- as.numeric(obsprob$z)

obsprob$z[obsprob$z>=62] = obsprob$z[obsprob$z>=62]+1
obsprob$z = 0.04 + (obsprob$z-1)*0.01

obsprob = obsprob[order( obsprob[,2], obsprob[,1] ),]
probabilities = list(x = seq(5,295,10), y = seq(5,295,10), z = matrix(obsprob$z, ncol = 30))
image.plot(probabilities, main = "Pine tree observation probabilities")

#interpolation <- interp(obsprob$x, obsprob$y, obsprob$z)
#image.plot(interpolation)

obspines$z = obspines$z - 1
obspines = obspines[order( obspines[,2], obspines[,1] ),]
trees = list(x = seq(5,295,10), y = seq(5,295,10), z = matrix(obspines$z, ncol = 30))

image.plot(trees, main = "Observations of pine trees")
```

We further assume that the observations given the true number of trees are spatially uncorrelated from one grid unit to another. We then specify the likelihood model for the observations. For each grid unit $i$ one has a binomial likelihood model with $d_i \leq k_i$,
$$[d_i|k_i] \sim p(d_i|k_i) = \begin{pmatrix}k_i\\d_i \end{pmatrix} \alpha_i^{d_i}[1-\alpha_i]^{k_i-d_i}.$$
Because the observations are conditionally independent from one node to another, the likelihood model with $\mathbf{d}\leq \mathbf{k}$ is,
$$[\mathbf{d}|\mathbf{k}] \sim p(\mathbf{d}|\mathbf{k}) = \prod_{i=1}^n p(d_i|k_i) = \prod_{i=1}^n \begin{pmatrix} k_i\\d_i\end{pmatrix} \alpha_i^{d_i}[1-\alpha_i]^{k_i-d_i}.$$

## b)
To further specify our model, we assume apriori that the distribution of trees is according to a stationary Poisson RF with model parameter $\lambda_k$. The prior Poisson model is defined on the discretized event count representation $\mathbf{k}\in \mathbb{N}^n_\oplus$ as follows,
$$ \mathbf{k} \sim p(\mathbf{k}) = \prod_{i=1}^n p(k_i) = \prod_{i=1}^n \frac{\pi_k^{k_i}}{k_i!}\times \exp\{-\pi_k\} =  \frac{\pi_k^{\sum_{i}k_i}}{\prod_{i}k_i!}\times \exp\{-n\pi_k\},$$
which is a discretized spatially stationary Poisson count RF with $\{\lambda(\mathbf{x}) = \lambda_k;\mathbf{x}\in D\}$ and $\pi_k = \lambda_k \Delta_n$ and basis-event set $\mathbb{E} = \emptyset$.

## c)
Now we want to estimate the intensity $\lambda_k$ based on the observations with observation probabilities. To estimate the intensity, we start by estimating the event probability $\pi_k$ in the discretized field represented by the $n$-vector $\mathbf{k}$. Hence, the model parameter for $\mathbf{k}$ is $\theta_{pP} = (\pi_k,\emptyset)$. This is assessed through marginal maximum likelihood estimation,
$$\hat \theta_{pP} = \underset{\theta_p}{\operatorname{arg max}}\{\log p(\mathbf{d};\theta_p)\}.$$
By utilizing the joint density of $\mathbf{k}$ and $\mathbf{d}$, and summing each term over $k_i\geq d_i$, we obtain the solution,
$$\hat \pi_k = \frac{n_d}{\boldsymbol \alpha^T \mathbf{i}_n},$$
where $n_d$ is the number of event observations and $\boldsymbol \alpha$ is the observation probabilities. Further, we get the estimated intensity $\hat \lambda_k = \hat \pi_k/\Delta_n$.

Having estimated the intensity, we generate 10 realizations from the prior Poisson event-count model and the associated approximate Poisson event-location realizations. The event-locations are then displayed.
```{r, echo =F, eval = T, out.width = "33%"}
pi_k = sum(obspines$z)/sum(obsprob$z)
lambda_k = pi_k/100

n = length(obspines$z)
k <- rpois(n,pi_k)
trees2 = list(x = seq(5,295,10), y = seq(5,295,10), z = matrix(k, ncol = 30))
image.plot(trees2, main = "Realization of prior distribution")

k <- rpois(n,(1-obsprob$z)*pi_k)+obspines$z
trees3 = list(x = seq(5,295,10), y = seq(5,295,10), z = matrix(k, ncol = 30))
image.plot(trees3, main = "Realization of posterior distribution")
image.plot(trees, main = "Observations of pine trees")
```

```{r, echo = F, eval = T, out.width = "20%"}
generate_event_locations <- function(pi_k,k_zero){
  k <- rpois(n,pi_k)+k_zero
  event_locations <- matrix(NA, ncol = 2, nrow = sum(k))
  count = 1
  for (i in 1:900){
    num = k[i]
    if (num!=0){
      x = runif(num,0,10)
      y = runif(num,0,10)
      event_locations[(count):(count+num-1),1] = x + ((i-1)%%30)*10
      event_locations[(count):(count+num-1),2] = y + ((i-1)%/%30)*10
      }
    count = count + num
    }
  return(event_locations)
  }
x = list()
for (i in 1:10){
  x[[i]] <- generate_event_locations(pi_k,rep(0,n))
  colnames(x[[i]]) = c("x","y")
}
p1 <- ggplot(as.data.frame(x[[1]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 1")
p2 <- ggplot(as.data.frame(x[[2]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 2")
p3 <- ggplot(as.data.frame(x[[3]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 3")
p4 <- ggplot(as.data.frame(x[[4]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 4")
p5 <- ggplot(as.data.frame(x[[5]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 5")
p6 <- ggplot(as.data.frame(x[[6]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 6")
p7 <- ggplot(as.data.frame(x[[7]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 7")
p8 <- ggplot(as.data.frame(x[[8]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 8")
p9 <- ggplot(as.data.frame(x[[9]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 9")
p10 <- ggplot(as.data.frame(x[[10]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 10")
p1
p2
p3
p4
p5
p6
p7
p8
p9
p10
```

## d)
Having defined the likelihood model and the prior distribution, we now want to develop the expression for the posterior discretized event-count model. To do this, we first write the joint distribution for the spatial variable $\mathbf{k}$ and the observations $\mathbf{d}$,
\begin{align*}
\begin{bmatrix} \mathbf{k}\\ \mathbf{d} \end{bmatrix} \sim p(\mathbf{k},\mathbf{d}) &= p(\mathbf{d}|\mathbf{k})p(\mathbf{k}) = \prod_{i=1}^n p(k_i|d_i)\prod_{i=1}^n p(k_i) = \prod_{i=1}^n p(d_i,k_i)\\
&= \prod_{i=1}^n \begin{pmatrix} k_i\\ d_i \end{pmatrix} \alpha_i^{d_i}[1-\alpha_i]^{k_i-d_i}\times \frac{\pi_k^{k_i}}{k_i!} \exp\{-\pi_k\}\\
&= \prod_{i=1}^n \frac{[\alpha_i\pi_k]^{d_i}}{d_i!}\times \frac{[(1-\alpha_i)\pi_k]^{k_i-d_i}}{(k_i-d_i)!} \times \exp\{-\pi_k\}. 
\end{align*}
Having arrived at the joint distribution, we note that it is on factorial form, which implies independence of each grid node. Hence, the normalizing constant can be found by summing for each node over $k_i\geq d_i$ to obtain $p(d_i)$. This means that the posterior distribution can be written as,
\begin{align*}
[\mathbf{k}|\mathbf{d}] \sim p(\mathbf{k}|\mathbf{d}) &= \prod_{i=1}^n [p(d_i)]^{-1}\times p(d_i,k_i) = \prod_{i=1}^n p(k_i|d_i)\\
&= \prod_{i=1}^n \frac{[(1-\alpha_i)\pi_k]^{k_i-d_i}}{(k_i-d_i)!} \times \exp\{-(1-\alpha_i)\pi_k\},
\end{align*}
for $\mathbf{k}\in \mathbb{N}_\oplus^n;\mathbf{k}\geq\mathbf{d}$. This is a discretized Poisson count random field with model parameters $\mathbf{\pi}_n = \Big((1-\alpha_1)\pi_k, (1-\alpha_2)\pi_k,\dots,(1-\alpha_n)\pi_k\Big)$ and base event set $\mathbf{k}_n^0 = (d_1,\dots,d_n)$.

To simulatefrom the associated approximate event-location model, we use Algorithm 7 from [@Omre].
```{r, echo = F, eval = T, out.width = "20%"}
x = list()
for (i in 1:10){
  x[[i]] <- generate_event_locations((1-obsprob$z)*pi_k, obspines$z)
  colnames(x[[i]]) = c("x","y")
}

p1 <- ggplot(as.data.frame(x[[1]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 1")
p2 <- ggplot(as.data.frame(x[[2]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 2")
p3 <- ggplot(as.data.frame(x[[3]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 3")
p4 <- ggplot(as.data.frame(x[[4]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 4")
p5 <- ggplot(as.data.frame(x[[5]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 5")
p6 <- ggplot(as.data.frame(x[[6]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 6")
p7 <- ggplot(as.data.frame(x[[7]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 7")
p8 <- ggplot(as.data.frame(x[[8]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 8")
p9 <- ggplot(as.data.frame(x[[9]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 9")
p10 <- ggplot(as.data.frame(x[[10]]), aes(x, y)) + geom_point()+  theme(legend.position = "none",axis.title.x=element_blank(),
        axis.text.y=element_blank(),axis.title.y=element_blank(), plot.title = element_text(size=22))  +
    ggtitle("Realization 10")
p1
p2
p3
p4
p5
p6
p7
p8
p9
p10
```


## Problem 3: Clustered event spatial variables

The Neyman-Scott cluster model for event spatial variables is 

$$N(x) \sim Poisson(\lambda(x)) \\$$